The benefits of reporting critical effect size values

### 2024-10-01 paper/paper.Rmd
^^

abstract: |
  Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size. These values are highly applicable in various scenarios, as they do not require specifying a plausible effect size of interest. Reporting critical effect size values may be especially useful when adopting NHST (Null Hypothesis Significance Testing) and the sample size has not been planned a priori, or when there is uncertainty about the expected effect size. To assist researchers in calculating critical effect size values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations.

@@

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference using statistical significance. The "critical effect size value(s)" refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.

>>>>

abstract: |
  Critical effect size values represent the smallest detectable effect that would reach statistical significance given a certain sample size and alpha level. It can be useful to calculate the smallest effect that can be statistically significant and evaluate whether such effects are plausible, and if desireable that smaller effects can be detected. Reporting critical effect size values may be especially useful while adopting NHST (Null Hypothesis Significance Testing) when the sample size has not been planned a priori, or when there is uncertainty about the expected effect size that can be collected. To assist researchers in calculating critical effect size values, we have developed an R package that allows to report the critical effect size values of model parameters. The package's functions are implemented for group comparisons, correlations, linear regression, and meta-analysis. What we propose could benefit researchers during the planning phase of the study by helping them to understand the limitations and strengths of their research design, and also as a retrospective tool to potentially reframe original interpretations, especially non significant results.

@@

In the present paper, we suggest that researchers should systematically report critical effect size values when they make inference use significance levels in hypothesis tests to make claims about the presence or absence of effects. The "critical effect size value(s)" refers to the smallest statistically significant effect sizes, which depend on the test performed and the characteristics of the sample [@lakens2022sample]. To cut to the chase, let us consider a study that tests (using for example a two-tailed test) a bivariate correlation with $n = 20$ participants and uses $p < 0.05$ as the threshold for statistical significance: in this case, the smallest significant effects (critical effect size values) are Pearson's $r = -0.44$ and $r = 0.44$. As we will argue, critical effect size values can serve as a useful complement to aid in the interpretation of study results, to communicate their relevance, and to help assess their replicability, particularly when statistical significance is relied upon for inference. This is especially important in cases where the sample size was not (or could not) be predetermined based on a theory-driven target effect size, or when statistical power may be difficult to determine or deviates from an optimal level.
